simulation for the sparse factor loadings

### 2018-01-25 analysis/simulation3.Rmd
Files commited by wflow_commit.

---
title: "simulation for the sparse factor loadings"
author: "Wei Wang"
date: 2017-04-04
output: html_document
---

@@

```{r}

CVPMD=function(Y,c_s,K){
  N = dim(Y)[1]
  P = dim(Y)[2]
  colindex = matrix(sample(P,P),ncol = 5)
  rowindex = matrix(sample(N,N),ncol = 5)

  missing= array(0,dim = c(5,N,P))
  foldindex = array(0,dim = c(5,5,2))
  for(i in 1:5){
    for(j in 1:5){
      foldindex[i,j,1] = i
      foldindex[i,j,2] = (i+j) %% 5
    }
  }
  foldindex[which(foldindex == 0)] = 5
  for(i in 1:5){
    missing[i, , ] = Y
    for(j in 1:5){
      missing[i,rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]] = NA
    }
  }
  n_s = length(c_s)
  CVRMSE = rep(NA,n_s)
  minrmse = Inf
  opt_s = 0
  # here we only need one layer
  for(t_s in 1:n_s){
    rmse = rep(0,5)
    for(i in 1:5){
      out = PMD(missing[i,,], sumabs = c_s[t_s], sumabsv = NULL, sumabsu = NULL,K = K)
      misshat = out$u %*%  diag(out$d) %*% t(out$v)
      for(j in 1:5){
        rmse[i] = rmse[i] + sum((Y[rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]] - 
                                   misshat[rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]])^2)
      }
      rmse[i] = sqrt(rmse[i] / (N * P/5))
    }
    CVRMSE[t_s] = mean(rmse)
    if(CVRMSE[t_s] < minrmse){
      minrmse = CVRMSE[t_s]
      opt_s = c_s[t_s]
    }
  }

  return(list(opt_s = opt_s ))
}

PMA.wrapper = function(Y,ngrids = 10,K){
  library(PMA)
  N = dim(Y)[1]
  P = dim(Y)[2]
  c_s = seq(0.2,0.8,len=ngrids)
  cvout = CVPMD(Y,c_s, K = K)
  out = PMD(Y,sumabsu = NULL, sumabsv = NULL, sumabs = cvout$opt_s ,K = K)
  residual_PMD = Y -  out$u %*%  diag(out$d) %*% t(out$v)
  return(list(d = out$d, u = out$u, v = out$v, residual = residual_PMD))
}

```

```{r}
set.seed(99)
N = 150
P = 240
K = 3
L_true = array(0,dim = c(N,K))
F_true = array(0,dim = c(P,K))
F_true[1:80,1] = rnorm(80,0,0.5)
F_true[81:160,2] = rnorm(80,0,1)
F_true[161:240,3] = rnorm(80,0,2)
L_true[1:10,1] = rnorm(10,0,2)
L_true[11:60,2] = rnorm(50,0,1)
L_true[61:150,3] = rnorm(90,0,0.5)
G = L_true %*% t(F_true)
# generate Y
E = matrix(rnorm(N*P,0,2),nrow=N)
Y = L_true %*% t(F_true) + E
g_flash = flashr::greedy(Y,K = 5)
g_b_flash = flashr::backfitting(Y,initial_list = g_flash)

outcv = PMA.wrapper(Y,10,K = 3)
```

```{r}
par(mfrow = c(2,2))
image( t(apply(t(g_b_flash$l %*% t(g_b_flash$f)),2,rev)),col = grey(seq(0,1,length.out = 5)),main = "FLASH")
image(t(apply(t(outcv$u %*% diag(outcv$d) %*% t(outcv$v)),2,rev)),col = grey(seq(0,1,length.out = 5)),main = "PMD")
image(t(apply(t(Y),2,rev)),col = grey(seq(0,1,length.out = 5)),main = "observed data")
image(t(apply(t(G),2,rev)),col = grey(seq(0,1,length.out = 5)),main = "underlying structure")
par(mfrow = c(2,2))
image( t(apply(t(g_b_flash$l %*% t(g_b_flash$f)),2,rev)),col =blues9,main = "FLASH")
image(t(apply(t(outcv$u %*% diag(outcv$d) %*% t(outcv$v)),2,rev)),col = blues9,main = "PMD")
image(t(apply(t(Y),2,rev)),col = blues9,main = "observed data")
image(t(apply(t(G),2,rev)),col = blues9,main = "underlying structure")
```

```{r}
library(ggplot2)
library(cowplot)
mat = cbind(L_true,g_b_flash$l,outcv$u)
mat[,8] = -mat[,8]
mat[,9] = -mat[,9]
mat[,6] = -mat[,6]
mat1 = sapply(seq(1,dim(mat)[2]), function(x){ mat[,x]/sqrt(sum((mat[,x])^2)) } )
colnames(mat1) = c("truth.l1","truth.l2","truth.l3","FLASH.l3","FLASH.l2","FLASH.l1","PMD.u3","PMD.u2","PMD.u1")
df1 = data.frame(mat1)
p1_1_l <- ggplot(df1, aes(truth.l1, FLASH.l1)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 1 (FLASH)")
p1_2_l <- ggplot(df1, aes(truth.l2, FLASH.l2)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 2 (FLASH)")
p1_3_l <- ggplot(df1, aes(truth.l3, FLASH.l3)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 3 (FLASH)")
p2_1_l <- ggplot(df1, aes(truth.l1, PMD.u1)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 1 (PMD)")
p2_2_l <- ggplot(df1, aes(truth.l2, PMD.u2)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 2 (PMD)")
p2_3_l <- ggplot(df1, aes(truth.l3, PMD.u3)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("loading 3 (PMD)")

library(ggplot2)
library(cowplot)
mat = cbind(F_true,g_b_flash$f,outcv$v)
mat[,8] = -mat[,8]
mat[,9] = -mat[,9]
mat[,6] = -mat[,6]
mat1 = sapply(seq(1,dim(mat)[2]), function(x){ mat[,x]/sqrt(sum((mat[,x])^2)) } )
colnames(mat1) = c("truth.f1","truth.f2","truth.f3","FLASH.f3","FLASH.f2","FLASH.f1","PMD.v3","PMD.v2","PMD.v1")
df1 = data.frame(mat1)
p1_1 <- ggplot(df1, aes(truth.f1, FLASH.f1)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 1 (FLASH)")
p1_2 <- ggplot(df1, aes(truth.f2, FLASH.f2)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 2 (FLASH)")
p1_3 <- ggplot(df1, aes(truth.f3, FLASH.f3)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 3 (FLASH)")
p2_1 <- ggplot(df1, aes(truth.f1, PMD.v1)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 1 (PMD)")
p2_2 <- ggplot(df1, aes(truth.f2, PMD.v2)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 2 (PMD)")
p2_3 <- ggplot(df1, aes(truth.f3, PMD.v3)) + geom_point() +  geom_abline(intercept = 0, slope = 1)+theme_gray() + ggtitle("factor 3 (PMD)")
plot_grid(p1_1_l, p2_1_l, p1_1, p2_1,
          p1_2_l, p2_2_l, p1_2, p2_2,
          p1_3_l, p2_3_l, p1_3, p2_3,labels = "" , ncol = 4)
```

>>>>

---
title: "simulation for the sparse factor loadings"
author: "Wei Wang"
date: 2017-04-04
output: 
  html_document:
    code_folding: hide
---

@@

This file is just generate the structure heatmap of rank 3 data.

```{r}
DataMaker = function(sigma = 2,seed_num = 99){
  set.seed(seed_num)
  N = 150
  P = 240
  K = 3
  L_true = array(0,dim = c(N,K))
  F_true = array(0,dim = c(P,K))
  F_true[1:80,1] = rnorm(80,0,0.5)
  F_true[81:160,2] = rnorm(80,0,1)
  F_true[161:240,3] = rnorm(80,0,2)
  L_true[1:10,1] = rnorm(10,0,2)
  L_true[11:60,2] = rnorm(50,0,1)
  L_true[61:150,3] = rnorm(90,0,0.5)
  G = L_true %*% t(F_true)
  # generate Y
  E = matrix(rnorm(N*P,0,sigma),nrow=N)
  Y = L_true %*% t(F_true) + E
  return(list(Y = Y, L_true = L_true, F_true = F_true, E = E))
}

CVPMD_softImpute=function(Y,c_s,K,fold = 10, method = "PMD"){
  N = dim(Y)[1]
  P = dim(Y)[2]
  colindex = matrix(sample(P,P),ncol = fold)
  rowindex = matrix(sample(N,N),ncol = fold)

  missing= array(0,dim = c(fold,N,P))
  foldindex = array(0,dim = c(fold,fold,2))
  for(i in 1:fold){
    for(j in 1:fold){
      foldindex[i,j,1] = i
      foldindex[i,j,2] = (i+j) %% fold
    }
  }
  foldindex[which(foldindex == 0)] = fold
  for(i in 1:fold){
    missing[i, , ] = Y
    for(j in 1:fold){
      missing[i,rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]] = NA
    }
    missing[i,,which(colSums(missing[i,,],na.rm = T) ==0)] = Y[,which(colSums(missing[i,,],na.rm = T) ==0)]
  }
  # c_s is the candicate of shrinkage parameter
  n_s = length(c_s)
  # rmse for each grids
  CVRMSE = rep(0,n_s)
  minrmse = Inf
  opt_s = 0
  # for each candidate, we run it N_sim times
  for(t_s in 1:n_s){
    # for each grid
    # each time we set the rmse to zeros
    rmse = rep(0,fold)
    for(i in 1:fold){
      if(method == "PMD"){
        res_log = capture.output({out = PMD(missing[i,,], sumabs = c_s[t_s], sumabsv = NULL, sumabsu = NULL,K = K)})
      }else{
        out = softImpute(missing[i,,], rank.max = K,lambda = c_s[t_s])
      }
      if(length(out$d)==1){
        misshat = (out$d) * out$u %*% t(out$v)
      }else{
        misshat = out$u %*%  diag(out$d) %*% t(out$v)
      }
      for(j in 1:fold){
        # for each fold j
        rmse[i] = rmse[i] + sum((Y[rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]] -
                                   misshat[rowindex[,foldindex[j,i,1]],colindex[,foldindex[j,i,2]]])^2,na.rm = TRUE)
      }
    } #get the result for one run
    CVRMSE[t_s] = CVRMSE[t_s] + sqrt(sum(rmse)/(N*P))
    if(CVRMSE[t_s] < minrmse){
      minrmse = CVRMSE[t_s]
      opt_s = c_s[t_s]
    }
  }
  return(list(opt_s = opt_s, output = CVRMSE))
}

PMA.wrapper = function(Y_data, ngrids = 10, K=3, fold = 10){
  library(PMA)
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  c_s = seq(0.1,0.9,len=ngrids)
  cvout = CVPMD_softImpute(Y,c_s,K ,fold , method = "PMD")
  res_log = capture.output({out = PMD(Y,sumabsu = NULL, sumabsv = NULL, sumabs = cvout$opt_s ,K = K)})
  Y_hat = out$u %*% diag(out$d) %*%  t(out$v)
  return(Y_hat)
}

softImpute.wrapper = function(Y_data, ngrids = 10, K = 3, fold = 10){
  library(softImpute)
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  c_s = seq(0,100,len=ngrids)
  cvout = CVPMD_softImpute(Y,c_s,K ,fold , method = "softImpute")
  out = softImpute(Y, rank.max = K,lambda = cvout$opt_s)
  if(length(out$d)==1){
    Y_hat = (out$d) * out$u %*% t(out$v)
  }else{
    Y_hat = out$u %*%  diag(out$d) %*% t(out$v)
  }
  return(Y_hat)
}

SVD.wrapper = function(Y_data,K = 3){
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  gsvd = svd(Y,nu = 3,nv = 3)
  Y_hat = (gsvd$u[,1:K] %*% diag(gsvd$d[1:K]) %*% t(gsvd$v[,1:K]))
  return(Y_hat)
}

SSVD.wrapper = function(Y_data,K = 3){
  library(ssvd)
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  gssvd = ssvd::ssvd(Y,method = "method",r = K)
  Y_hat = (gssvd$u %*% diag(gssvd$d) %*% t(gssvd$v))
  return(Y_hat)
}

flashBF.wrapper = function(Y_data,K = 3){
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  library(ebnm)
  library(flashr2)
  data =  flashr2::flash_set_data(Y)
  f_greedy = flashr2::flash_add_greedy(data,verbose=F,var_type = "constant", K = K)
  f = flashr2::flash_backfit(data,f_greedy,var_type = "constant")
  Y_hat = f$EL %*% t(f$EF)
  return(Y_hat)
}

flash.wrapper = function(Y_data, K = 3){
  # missindex is a matirx with 3 column here: i j x
  # Y has miss value already
  Y = Y_data$Y
  L_true = Y_data$L_true
  F_true = Y_data$F_true
  E = Y_data$E
  N = dim(Y)[1]
  P = dim(Y)[2]
  data =  flashr2::flash_set_data(Y)
  g_flash = flashr2::flash_add_greedy(data,verbose=F,var_type = "constant", K = K)
  Y_hat = g_flash$EL %*% t(g_flash$EF)
  return(Y_hat)
}
```

```{r}
library(PMA)
library(flashr2)
library(ssvd)
library(softImpute)
Y_data = DataMaker(seed_num = 0)

LF_1 = PMA.wrapper(Y_data)
LF_2 = flash.wrapper(Y_data)
LF_3 = flashBF.wrapper(Y_data)
LF_4 = SVD.wrapper(Y_data)
LF_5 = SSVD.wrapper(Y_data)
LF_6 = softImpute.wrapper(Y_data)
LF_true = Y_data$L_true %*% t(Y_data$F_true)

library(reshape2)
melted_cormat_1 <- melt(LF_1, na.rm = TRUE)
melted_cormat_2 <- melt(LF_2, na.rm = TRUE)
melted_cormat_3 <- melt(LF_3, na.rm = TRUE)
melted_cormat_4 <- melt(LF_4, na.rm = TRUE)
melted_cormat_5 <- melt(LF_5, na.rm = TRUE)
melted_cormat_6 <- melt(LF_6, na.rm = TRUE)
melted_cormat_true <- melt(LF_true, na.rm = TRUE)
# Heatmap
library(ggplot2)
plot_HP = function(title_name = "truth", melted_cormat){
  p1 = ggplot(data = melted_cormat, aes(Var2, Var1, fill = value))+
   geom_tile(color = "white")+
    scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
     midpoint = 0, limit = c(-2,2), space = "Lab") + labs(title = title_name, y = "samples", x = "variables") +
    theme_minimal() + theme(legend.position="none") 
  p1
}

p0 = plot_HP(title_name = "Truth", melted_cormat = melted_cormat_true)
p1 = plot_HP(title_name = "PMD.cv1", melted_cormat = melted_cormat_1)
p2 = plot_HP(title_name = "flash", melted_cormat = melted_cormat_2)
p3 = plot_HP(title_name = "flashb", melted_cormat = melted_cormat_3)
p4 = plot_HP(title_name = "SVD", melted_cormat = melted_cormat_4)
p5 = plot_HP(title_name = "SSVD", melted_cormat = melted_cormat_5)
p6 = plot_HP(title_name = "SI.cv", melted_cormat = melted_cormat_6)
```

```{r}
gridExtra::grid.arrange(p0,p5,p6,p4,p1,p2,ncol = 2)
```
